---
title: "Breast Cancer Project"
author: "Nicholas Kaufman"
date: "2025-12-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set((echo = FALSE), warning = FALSE, 
root.dir = "Users/nickkaufman/Recents")
library(readr)
library(dplyr)
library(tidyr)
library(caret)
library(class)
library(glmnet)
library(mgcv)
library(corrplot)
library(pROC)
library(ggplot2)
```

## Data Preparation
```{r, message=FALSE, warning=FALSE}
cancer_raw <- read_csv("Downloads/data.csv", show_col_types = FALSE)

cancer <- cancer_raw %>%
  janitor::clean_names() %>%
  select(diagnosis:fractal_dimension_worst) %>%
  mutate(
    diagnosis = factor(diagnosis,
                       levels = c("B", "M"),
                       labels = c("Benign", "Malignant"))
  )

head(cancer, 5)

table(cancer$diagnosis)

```
The dataset used in this analysis contains measurements of tumor characteristics 
derived from digitized images of breast masses. The objective is to use these 
features to classify tumors as either benign or malignant.

The data was first cleaned and prepared by standardizing column names and 
converting the diagnosis variable into a factor with clearly defined classes. 
This ensures consistency in modeling and interpretation throughout the analysis.


## Exploratory Data Analysis
```{r}

eda_means <- cancer %>%
  group_by(diagnosis) %>%
  summarize(
    radius_mean = mean(radius_mean, na.rm = TRUE),
    perimeter_mean = mean(perimeter_mean, na.rm = TRUE),
    smoothness_mean = mean(smoothness_mean, na.rm = TRUE),
    compactness_mean = mean(compactness_mean, na.rm = TRUE),
    concavity_mean = mean(concavity_mean, na.rm = TRUE),
    concave_points_mean = mean(concave_points_mean, na.rm = TRUE),
    symmetry_mean = mean(symmetry_mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(-diagnosis, names_to = "variable", values_to = "mean_value")

# Faceted bar plot of mean values
mean_plot <- ggplot(eda_means, aes(x = diagnosis, y = mean_value, 
fill = diagnosis)) +
  geom_col(position = "dodge") +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Mean Feature Values by Tumor Diagnosis",
    x = "Diagnosis",
    y = "Mean Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(mean_plot)



```


Initial exploratory analysis showed that malignant tumors tended to have higher 
average values across several features, including radius, perimeter, 
compactness, concavity, concave points, and symmetry. These differences 
suggested that the predictor variables may carry meaningful information for 
distinguishing benign from malignant tumors.


```{r, fig.width=10, fig.height=8}
numeric_data <- cancer %>%
  mutate(diagnosis = ifelse(diagnosis == "Malignant", 1, 0))

cor_target <- cor(numeric_data, use = "complete.obs")[, "diagnosis"]

cor_df <- data.frame(
  variable = names(cor_target),
  correlation = cor_target
) %>%
  filter(variable != "diagnosis") %>%
  arrange(desc(abs(correlation)))

ggplot(cor_df, aes(x = reorder(variable, correlation), y = correlation)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Correlation of Features with Tumor Diagnosis",
    x = "Feature",
    y = "Correlation"
  ) +
  theme_minimal()


```


Correlation analysis further showed that many size-related variables, including 
radius, perimeter, and area, were highly correlated with one another. This 
indicates a significant degree of redundancy within the feature set.

In contrast, features related to tumor shape and boundary irregularity, such as 
concavity and concave points, also demonstrated strong relationships with 
diagnosis while being less redundant.

These findings suggest that a smaller subset of variables may be sufficient for accurate classification, and that feature selection could improve model stability.


```{r}
top_vars <- cor_df %>%
slice_max(abs(correlation), n = 3) %>%
pull(variable)

plot_data <- cancer %>%
	pivot_longer(
		cols = all_of(top_vars),
		names_to = "variable",
		values_to = "value"
	)
ggplot(plot_data, aes(x = value, fill = diagnosis)) +
	geom_density(alpha = 0.3) +
	facet_wrap(~ variable, scales = "free", ncol = 1) +
	theme_minimal()

ggplot(plot_data, aes(x = diagnosis, y = value, fill = diagnosis)) +
	geom_boxplot() +
	facet_wrap(~ variable, scales = "free") +
	theme_minimal() +
	theme(legend.position = "none")
```


Distribution plots of the most strongly correlated variables showed clear 
separation between benign and malignant tumors. In particular, features such as 
concave points and radius displayed distinct distributions across the two 
classes, reinforcing their importance as predictive variables.

Overall, the exploratory analysis indicated that the dataset contains strong 
predictive signals, but also highlighted the presence of highly correlated 
features. These observations informed the modeling approach, including the use 
of regularization techniques to reduce redundancy and improve model 
interpretability.

## Logistic Regression Model
```{r}
set.seed(123)



train_index <- caret::createDataPartition(cancer$diagnosis, p = 0.8, 
list = FALSE)
train <- cancer[train_index, ]
test  <- cancer[-train_index, ]

model <- glm(diagnosis ~ ., data = train, family = binomial)


probs <- predict(model, newdata = test, type = "response")


preds <- ifelse(probs > 0.5, "Malignant", "Benign")
preds <- factor(preds, levels = levels(test$diagnosis))

cat("=== Logistic Regression Results ===\n")
caret::confusionMatrix(preds, test$diagnosis, positive = "Malignant")

# ROC and AUC
roc_obj <- pROC::roc(test$diagnosis, probs)

cat("\n=== ROC AUC ===\n")
pROC::auc(roc_obj)

plot(roc_obj)

```


A logistic regression model was used as a baseline approach to classify tumors 
as benign or malignant. The model was trained on 80% of the dataset and 
evaluated on the remaining 20%.

The model achieved strong performance on the test set, with high accuracy, 
sensitivity, and specificity, along with an AUC of approximately 0.96. This 
indicates that the predictor variables contain substantial information for 
distinguishing between benign and malignant tumors.

While the logistic regression model performed well, the presence of highly 
correlated predictors suggested potential redundancy in the feature set. This 
motivated the use of a regularized modeling approach to improve interpretability 
and reduce the number of predictors.

## Model Validation
```{r}

	set.seed(123)
	
	control <- caret::trainControl(
	  method = "cv",
	  number = 5,
	  classProbs = TRUE,
	  summaryFunction = twoClassSummary,
	  savePredictions = TRUE
	)
	
	cv_model <- caret::train(
	  diagnosis ~ .,
	  data = cancer,
	  method = "glm",
	  family = "binomial",
	  metric = "ROC",
	  trControl = control
	)
	
	print(cv_model)


```

While the model performed well, the presence of highly correlated predictors 
suggested redundancy in the feature set. This motivated the use of 
regularization to improve interpretability and reduce the number of predictors.

## Regularized Logistic Regression
```{r}

x <- model.matrix(diagnosis ~ ., cancer)[, -1]
y <- cancer$diagnosis

set.seed(123)

cv_glmnet <- glmnet::cv.glmnet(
  x = x,
  y = y,
  family = "binomial",
  alpha = 1
)

plot(cv_glmnet)
print(cv_glmnet$lambda.min)
print(cv_glmnet$lambda.1se)


```
To reduce redundancy among predictors and improve model interpretability, a 
regularized logistic regression model using Lasso (L1) regularization was 
implemented.

Lasso regularization adds a penalty to the model that shrinks less important 
coefficients toward zero, effectively performing feature selection. This is 
particularly useful in this dataset, where many predictors are highly correlated.

The model was trained using cross-validation to identify the optimal level of 
regularization. The results showed that a range of regularization strengths 
produced similar performance, indicating that a simpler model could be used 
without sacrificing predictive accuracy.

Using the more regularized model (lambda.1se), the number of predictors was 
reduced substantially, while maintaining strong classification performance. 
This confirms that many of the original features were redundant and that a 
smaller subset of variables is sufficient for accurate prediction.

The features retained in the regularized model were primarily related to tumor 
shape and boundary irregularity, such as concave points, concavity, and 
symmetry. This suggests that morphological complexity plays a more important 
role in distinguishing malignant tumors than size alone.

```{r}

coef_matrix <- coef(cv_glmnet, s = "lambda.1se")

nonzero_coef <- coef_matrix[coef_matrix[,1] != 0, , drop = FALSE]

cat("=== Selected Features (Lasso - lambda.1se) ===\n")
print(nonzero_coef)
```

## Final Model Evaluation
```{r}

x_test <- model.matrix(diagnosis ~ ., test)[, -1]


probs_glmnet <- predict(cv_glmnet, newx = x_test, s = "lambda.1se", 
type = "response")


preds_glmnet <- ifelse(probs_glmnet > 0.5, "Malignant", "Benign")
preds_glmnet <- factor(preds_glmnet, levels = levels(test$diagnosis))


caret::confusionMatrix(preds_glmnet, test$diagnosis, positive = "Malignant")

```

The regularized logistic regression model was evaluated on the test set to 
assess its performance on unseen data.

The model achieved high classification accuracy, with strong sensitivity and 
specificity. Only a small number of observations were misclassified, indicating 
that the model performs reliably across both classes.

In addition to accuracy metrics, the model demonstrated near-perfect 
discrimination ability, with an area under the ROC curve (AUC) of 0.9993. 
This confirms that the model is highly effective at distinguishing between 
benign and malignant tumors.

Importantly, this level of performance was achieved using a reduced set of 
predictors, demonstrating that a simpler model can perform just as well as a 
more complex one.


```{r}

probs_glmnet <- as.numeric(probs_glmnet)
roc_obj_glmnet <- pROC::roc(test$diagnosis, probs_glmnet)
pROC::auc(roc_obj_glmnet)
plot(roc_obj_glmnet)

```
## Conclusion

This project demonstrated that breast cancer tumors can be accurately classified 
using a relatively small subset of features derived from tumor measurements. 
It is important to note that this dataset exhibits strong separation between 
benign and malignant tumors, which contributes to the high performance observed 
across models. As a result, the near-perfect classification metrics may not 
fully reflect performance on more complex or less separable real-world data.

While a standard logistic regression model achieved strong performance, the 
presence of highly correlated predictors highlighted redundancy in the feature 
set. Cross-validation confirmed that the predictive signal in the dataset was 
robust, while the correlated predictors reinforced the value of a regularized 
approach for feature selection and interpretability.

By applying Lasso regularization, the model was simplified significantly, 
reducing the number of predictors while maintaining high classification 
performance. The final model achieved strong accuracy, sensitivity, and 
specificity, along with near-perfect discrimination as measured by AUC.

Importantly, the results showed that features related to tumor shape and 
boundary irregularity played a more significant role in classification than 
size alone. This highlights the value of feature selection and regularization 
in building interpretable and reliable predictive models.
